Teaching (Digital) Epigraphy

### 2015-12-20 source/48/EAGLE16-lame-et-al-teaching.tex
Paper 48 (Lamé et al) added, but not checked

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

@@

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Teaching (Digital) Epigraphy}
\author[ilc]{Marion Lamé\corref{first}}
\author[zag]{Ivan Radman-Livaja}
\author[mau]{Bruce Robertson}
\author[isti]{Federico Ponchio}
\address[ilc]{Istituto di Linguistica Computazionale}
\address[isti]{Visual Computing Laboratory, ISTI-CNR}
\address[zag]{Archaeological Museum in Zagreb}
\address[mau]{Mount Allison University}
\cortext[first]{Corresponding author. Email: federico.ponchio@isti.cnr.it}

@@

\noindent Although undergraduate students are naturally attracted to inscriptions and epigraphy affords them an important window
on the past, they are seldom given the opportunity to study inscriptions directly. This is because epigraphical work is
seen as the specialist’s domain: the analysis of such material generally requires a rather high level of expertise,
normally acquired during graduate studies and beyond. Lacking sufficient skills and knowledge to comprehensively
understand often complex epigraphic data, undergraduate students are simply unable to offer the expert opinions
sometimes sought by a project manager. Thus, offering undergraduate students the opportunity to tackle tasks usually
reserved to their senior colleagues is certainly not a common occurrence. In our small project ``Pedagogical,
Scientific \& Technical Experiment in Digital Epigraphy: the Study Case of the Tesserarum Sisciae Sylloge (TSS) through
a Digital Autoptic Process (DAP)'' we aimed to challenge this state of affairs. Using digital methods and a
custom-designed course of rapid study, we offered undergraduates the opportunity to tackle epigraphic tasks usually
reserved to their senior colleagues. We made use of newly-digitized epigraphic material: Roman commercial lead tags
from the ancient city of Siscia.

@@

\noindent The Roman Department of the Archaeological Museum in Zagreb (AMZ) contains close to 1200 of these inscribed lead tags
found in Sisak (Siscia), one of the largest urban centers in south-western Pannonia. Most of them were found during the
dredging of the Kupa river before WWI. Since the dredging was localized in the very centre of the town, i.e. in front
of the old Roman port quarter, it would seem that all the tags come from a limited area. All of those tags are small
lead tablets, of a more or less rectangular shape, pierced with a hole so that the tag could be attached to the bags
containing the merchandise or to the merchandise itself with a small rope or a metal wire. They all carry an
inscription, most of the time on both sides. Those inscriptions are always, up to now, written in capital letters or
the older Roman cursive, sometimes even in a mixture of both. Most of the tags were reused several times and thus they
carry traces of older inscriptions (palimpsests). Those inscriptions usually follow the same model: one side of the tag
carries personal names, the other side carries an inscription mentioning the merchandise, most of the time in an
abbreviated form, as well as a price - normally expressed in denarii or fractions of the denarius - and quite often an
indication of quantity or weight.

@@

\begin{figure}
\centering
\includegraphics[width=2.944cm,height=2.752cm]{EAGLE16lameetalteaching-img001.jpg}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=2.944cm,height=2.82cm]{EAGLE16lameetalteaching-img002.jpg}
\end{figure}
Most of the tags, if not all, are linked to the wool trade and the textile industry. Words like lana, pan(n)um, tvnica,
sagvm, p(a)envla, pal(l)a, palliolum, lodix, banata, and abolla, appear regularly enough without being abbreviated and
thus the interpretation of common abbreviations like l, la, lan, pan, t, sag, paenv, pal, lo, lod, lodi, bana, ab is
not in doubt. The other abbreviations are mostly related to terms of colour. The prices on those tags are an important
information: they indicated the value of the goods or the cost of a given service (e.g. cleaning, fulling, dyeing). It
would also appear that they were used by fullers and dyers as ownership tags. By noting the name of the client as well
as the type of cloth or service and the price on the tag which was subsequently attached to the item to be processed
(i.e. cleaned or dyed), shop owners could easily return their property to the clients as well as charge them the
correct fee\footnote{\citep{radman_livaja_segestica..._2008}, \citep{radman-livaja_roetiquettes_2010}, \citep{radman-livaja_5._2013}, \citep[165-172]{radman-livaja_two_2013}, \citep{radman-livaja_plombs_2014},  for analogous tags with \citep[97-104]{mocsy_olom_1956}; \citep[195-210]{egger_funf_1968}; \citep[127-138]{frei-stolba_les_2011}; \citep[121-137/93, 215-222]{schwinden_romerzeitliche_1985}; \citep{romer-martijnse_romerzeitliche_1990}; \citep[5-48]{romer-martijnse_fruhkaiserzeitliche_1997}; \citep[301-305]{feugere_etiquette_1993}; \citep[211-220]{weiss_bleietiketten_1991}; \citep[29-40]{paci_etichette_1995}; \citep[207-216]{bassi_tre_1996}; \citep[121-135]{bizzarini_quattro_2005}; \citep[43-51]{buchi_etichette_2005}; \citep[42-110]{cresci_supellex_2010}; \citep[295-317]{jacques_artisanat_2010}; \citep[237-246]{wedenig_bleietikett_2013}.}.

@@

\noindent In fact, the transcription and interpretation of Roman inscribed lead tags are often challenging, and so we ignored how
far we could go with these students if useable results would ensue. First we made a selection of fourty tags to be
subjected to the adequate digitization process we chose to work with: Reflectance Transforming Imaging (RTI). It is a
small but nevertheless representative sample of the whole corpus. The students had at their disposal different kinds of
inscriptions, including both easily readable specimens and more challenging graffiti (e.g., No. 12582). One may also
profit from the assistance of undergraduate and graduate students, which could not often be the case
before\footnote{\ The Laboratorio di Cultura Digitale, lead by Prof. Salvatori, at the University of Pisa applies the
didactic model DIGICRAFT. This means that undergraduate students might share their skills, especially the digital one,
in order to allow an entire team to reach its goal. For more information see: http://www.labcd.unipi.it/laboratorio .}.
This experiment showed that their contribution can be useful, because technology may compensate for their lack of
knowledge although the final conclusions have to remain in the domain of scholars familiar with the subject. In any
case, this experiment shows how the digital edition may allow both scholars and students to have an open access to what
is basically a primary source of information for humanities (presented as a digital facsimile)\footnote{\ Terras, 2015
writes about some of the OA issues: ``While digitisation is not a pre-requisite to gaining access to material (...), and
while digital surrogates of cultural heritage objects do not have to be openly shared once created, just as the
sciences are calling for publication of source data as part of the Open Access movement, opening up access to primary
sources in the cultural heritage sector and encouraging them to be published in a way which is as accessible as
possible has the potential to change the nature of research outputs in the Humanities and Social Sciences, as well as
the nature of research itself in these areas.''}. Of course, learners could train themselves on the drawing of the tag,
but they would not confront the difficulty of interpreting the numerous ambiguities and the responsibility of deciding
what is on the tag.

@@

The drawing process consists in associating some heterogeneous elements of the wSystem and the tSystem. MarkOut is the
linking tool that allows to represent the graphico-textual (g/t) relationship between those two systems. MarkOut allows
the students to work on the wSystem of the inscriptions by drawing the signs over an image of the tag using the mouse.
The line is converted in a curve that can be later fixed using handles. Each mark, that represents a phenomenon of the
wSystem of the inscription, is assigned to some textual information such as a letter or a symbol (Unicode and/or XML
encoding), that represents a textual phenomenon of the tSystem of the inscription. This assignation creates inside the
MarkOut the g/t relationship used for query processing. The final result is an SVG file that can be easily parsed to
recover the linking information between wSystem and tSystem (e.g. browsing by glyph's shape and interpretation,
identifying glyph's position), and easily shared with anyone with a browser (e.g.: fig. 3.1). The linking tool MarkOut
is one of the editing components of the modular DAP. They fit into communication workflows\footnote{This raises, of
course, questions, most of them outside the purview of this article, about the relationship that evolves between
research and our modern societies. We invite the reader to peruse various initiatives such as Sabattini, 2006 on the «
valeur d'ancienneté » of cultural heritage, that participates to social well-being, the \ Civic Epistemologies
http://www.civic-epistemologies.eu and its Berlin charter, as well as Public History approaches
http://public-history-weekly.oldenbourg-verlag.de.} such as crowdsourcing, editing inscriptions, dialoguing with
experts, preserving and promoting cultural heritage and, \ as here, teaching some skills in epigraphy.

Fig. 3.1. MarkOut screenshots (detailed part of tag no. 12563) showing the association, expressed in red, of the Unicode
Latin characters with the position and the shape of the writing: U+0045 with II (identified as E2 by students) and
U+004C with L (L1 or L5).

\begin{figure}
\centering
\includegraphics[width=4.941cm,height=2.281cm]{EAGLE16lameetalteaching-img003.png}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=4.941cm,height=2.281cm]{EAGLE16lameetalteaching-img004.png}
\end{figure}
Students were involved for a very short period of time: on average two hours per day over a couple of days only (in
Spring 2014 and in Fall 2015). This training was delivered in three steps (``Readings'', ``Writings'', ``Test \& Taste a
DAP'') during which there were given some academic documentation about the lead tags, and basic knowledge about digital
epigraphy (epigraphic message in a computer, dispositive analysis, existing projects, digital tools, XML and Unicode
encoding, DAP, and issues regarding digital representation of artifacts).

Students encountered their first epigraphic challenges with tag no. 12582 (Radman-Livaja 2014: cat 03.13, 347), found in
the Kupa river in Sisak and offered by Andrija Colussi to the Archaeological Museum in 1898. It is a rather typical
lead tag of an irregular rectangular form, pierced with a hole. Its size (22.7x22.8x2.4 mm) corresponds fairly well to
the average size of other specimens found in Siscia. Although its surface is rather damaged and shows clear traces of
erasure, the most recent inscription remains quite readable. However, many traces of an older record (or several older
records?), complicate significantly the transcription and the interpretation. The last inscription can be read,
nevertheless, but it demands a certain effort and quite a lot of experience in Latin palaeography:

Obverse (palimpsest)\ \ \ \ Reverse\ \  \ \ \ \ \ \ \ \ \ \ \ \ (palimpsest)

\begin{figure}
\centering
\includegraphics[width=2.044cm,height=1.794cm]{EAGLE16lameetalteaching-img005.png}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=2.044cm,height=1.794cm]{EAGLE16lameetalteaching-img005.png}
\end{figure}
Ate[323?]r[323?]\ \  \ S[323?]i[323?]i[323?]x[323?]t[323?]i[323?]\ \  \ \ \ \ \ la(na) p(ondo) i\ \ . . . s

ivs\ \  \ i \ \ onis \ n\ \ \ \ cor(ticea)\ \ \ \ . . v\ \

\ \  \ i . n i\ \ \ \ \ \ [4FE?] =$-$ £\ \ \ \  \ iir

@@

\begin{figure}
\centering
\includegraphics[width=2.965cm,height=1.79cm]{EAGLE16lameetalteaching-img006.png}
\end{figure}

@@

\begin{figure}
\centering
 [Warning: Image ignored] % Unhandled or unsupported graphics:
%\includegraphics[width=7.329cm,height=4.404cm]{https://lh3.googleusercontent.com/HwLa5or_paTAyMS3Mu5XB3NucZXrDws6Jvc5WCC2ONHqBnBugWRtTF-YaKTLhJLhg6fShq3n_nG_37csBdKiKoSFA2tqVqhA0ovfyyT27h6KvGa-4slEmeURVSu_QG2AzjqTevFycw}

\end{figure}
No. 12563 is a rectangular lead tag of irregular form with a badly damaged and scratched surface. Nevertheless, the most
recent inscription is clearly readable while one may still discern traces of older inscriptions. An older inscription
might be visible on the obverse, perhaps series of numbers but its meaning remains unclear. Students have
transcriptions and drawings.

\begin{figure}
\centering
\includegraphics[width=2.845cm,height=1.896cm]{EAGLE16lameetalteaching-img007.png}
\end{figure}
Obverse\ \ \ \ \ \ Reverse

Aeli(i)\ \ \ \ \ \ \ \ l[323?]odicem

Tasti\ \ \ \ \ \ \ \ murtiolam[323?]

\ \ \ \ \ \ \ \ p(ondo) vi s

\ \ \ \ \ \ \ \ [4FE?] vi s

\begin{flushleft}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{|m{1.4399999cm}|m{0.518cm}|m{0.37700003cm}|m{0.361cm}|m{0.297cm}|m{0.361cm}|m{0.518cm}|m{0.486cm}|m{0.361cm}|m{0.30300003cm}|}
\hline
\centering 12563 (obv) &
A &
E &
L &
I &
T &
A &
S &
T &
I\\\hline
\centering Stud.1 &
A10 &
E2 &
L1 &
I1 &
T1 &
A10 &
S3 &
T2 &
I1\centering
\includegraphics[width=2.563cm,height=2.579cm]{EAGLE16lameetalteaching-img008.png}
\\\hline
\centering Stud. 2 &
A10 &
E2 &
L5 &
I1 &
T2 &
A10 &
S4? &
T2 &
I1\\\hline
\end{supertabular}
\end{flushleft}

\begin{figure}
\centering
\includegraphics[width=2.461cm,height=2.104cm]{EAGLE16lameetalteaching-img009.png}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=2.268cm,height=2.104cm]{EAGLE16lameetalteaching-img009.png}
\end{figure}

\begin{flushleft}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{|m{1.393cm}|m{0.361cm}|m{0.40900004cm}|m{0.40900004cm}|m{0.439cm}|m{0.393cm}|m{0.37700003cm}|m{0.444cm}|}
\hline
\centering 12563 (rev)\centering
\includegraphics[width=2.898cm,height=2.656cm]{EAGLE16lameetalteaching-img008.png}
 &
\centering L &
\centering O &
\centering D &
\centering I &
\centering C &
\centering E &
\centering\arraybslash M\\\hline
\centering Stud. 1 &
L2 &
O3 &
D3 &
I1 &
C2 &
E2 &
M5\\\hline
\centering Stud. 2 &
? &
O3 &
D5 &
I2? &
C2 &
E3 &
M5\\\hline
\end{supertabular}
\end{flushleft}

\begin{flushleft}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{|m{0.58cm}|m{0.37700003cm}|m{0.37700003cm}|m{0.361cm}|m{0.297cm}|m{0.40900004cm}|m{0.361cm}|m{0.518cm}|m{0.439cm}|m{0.37700003cm}|m{0.37700003cm}|m{0.297cm}|m{0.34900004cm}|}
\hline
\centering M &
\centering V &
\centering R &
\centering T &
\centering I &
\centering O &
\centering L &
\centering A &
\centering M &
\centering P &
\centering V &
\centering I &
\centering\arraybslash S\\\hline
M3 &
V3 &
R7 &
T2 &
I1 &
O3 &
L2 &
A10 &
M2 &
P1 &
V3 &
I1 &
S5\\\hline
M3? &
V4 &
R6 &
T2 &
I1 &
O3 &
L5 &
A10 &
M5 &
P2 &
V4 &
I1 &
S5\\\hline
\end{supertabular}
\end{flushleft}

@@

\begin{figure}
\centering
\includegraphics[width=2.009cm,height=1.96cm]{EAGLE16lameetalteaching-img010.png}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=2.103cm,height=1.961cm]{EAGLE16lameetalteaching-img010.png}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=2.044cm,height=1.794cm]{EAGLE16lameetalteaching-img011.png}
\end{figure}
Obverse \ \ \ (palimpsest)

Vale\ \  \ \ \ [4FE?] v[323?]i\ \

rius \ lis\ \  \ \ \ \ tiir \ r[323?] i

Martia\ \  \ \ \ \

@@

\begin{flushleft}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{|m{1.513cm}|m{0.37700003cm}|m{0.518cm}|m{0.361cm}|m{0.37700003cm}|m{0.37700003cm}|m{0.283cm}|m{0.37700003cm}|m{0.46900004cm}|m{0.439cm}|m{0.518cm}|m{0.518cm}|m{0.50200003cm}|m{0.297cm}|m{0.961cm}|}
\hline
\centering 13053 (obv.) &
\centering V &
\centering A &
\centering L &
\centering E &
\centering R &
\centering I &
\centering V &
\centering S &
\centering M &
\centering A &
\centering R &
\centering T &
\centering I &
\centering\arraybslash A\\\hline
\centering Stud. 1 &
V2 &
A10 &
L3 &
E3 &
R7 &
I2 &
V1 &
S6 &
M5 &
? &
R7 &
T3 &
? &
Ignored\\\hline
\centering Stud. 2 &
V2 &
A10 &
L3 &
E2 &
R7 &
I1 &
V1 &
S6? &
M5 &
A10 &
R6? &
T3? &
I1 &
A2\\\hline
\end{supertabular}
\end{flushleft}

>>>>

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}

\usepackage[normalem]{ulem}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\usepackage{teubner}
\languageattribute{greek}{polutoniko}

@@

% dummy text (remove in a normal thesis)
% remove if not necessary
%\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Teaching (Digital) Epigraphy}
\author[ilc]{Marion Lamé}
\author[zag]{Ivan Radman-Livaja}
\author[mau]{Bruce Robertson}
\author[isti]{Federico Ponchio\corref{first}}
\address[ilc]{Istituto di Linguistica Computazionale}
\address[isti]{Visual Computing Laboratory, ISTI-CNR}
\address[zag]{Archaeological Museum in Zagreb}
\address[mau]{Mount Allison University}
\cortext[first]{Corresponding author. Email: federico.ponchio@isti.cnr.it}

@@

\noindent Although undergraduate students are naturally attracted to inscriptions and epigraphy affords them an important window
on the past, they are seldom given the opportunity to study inscriptions directly. This is because epigraphical work is
seen as the specialist’s domain: the analysis of such material generally requires a rather high level of expertise,
normally acquired during graduate studies and beyond. Lacking sufficient skills and knowledge to comprehensively
understand often complex epigraphic data, undergraduate students are simply unable to offer the expert opinions
sometimes sought by a project manager. Thus, offering undergraduate students the opportunity to tackle tasks usually
reserved to their senior colleagues is certainly not a common occurrence. In our small project ``Pedagogical,
Scientific \& Technical Experiment in Digital Epigraphy: the Study Case of the \emph{Tesserarum Sisciae Sylloge} (TSS) through
a Digital Autoptic Process (DAP)'' we aimed to challenge this state of affairs. Using digital methods and a
custom-designed course of rapid study, we offered undergraduates the opportunity to tackle epigraphic tasks usually
reserved to their senior colleagues. We made use of newly-digitized epigraphic material: Roman commercial lead tags
from the ancient city of Siscia.

@@

\noindent The Roman Department of the Archaeological Museum in Zagreb (AMZ) contains close to 1200 of these inscribed lead tags
found in Sisak (Siscia), one of the largest urban centers in south-western Pannonia. Most of them were found during the
dredging of the Kupa river before WWI. Since the dredging was localized in the very centre of the town, i.e. in front
of the old Roman port quarter, it would seem that all the tags come from a limited area. All of those tags are small
lead tablets, of a more or less rectangular shape, pierced with a hole so that the tag could be attached to the bags
containing the merchandise or to the merchandise itself with a small rope or a metal wire. They all carry an
inscription, most of the time on both sides. Those inscriptions are always, up to now, written in capital letters or
the older Roman cursive, sometimes even in a mixture of both. Most of the tags were reused several times and thus they
carry traces of older inscriptions (palimpsests). Those inscriptions usually follow the same model: one side of the tag
carries personal names, the other side carries an inscription mentioning the merchandise, most of the time in an
abbreviated form, as well as a price - normally expressed in \emph{denarii} or fractions of the \emph{denarius} - and quite often an
indication of quantity or weight.

\begin{figure}[!h]
\centering
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img001.jpg}
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img002.jpg}
\caption{No. 12582.}
\label{fig:12582}
\end{figure}

@@

\begin{figure}
\centering

\end{figure}
Most of the tags, if not all, are linked to the wool trade and the textile industry. Words like \emph{LANA, PAN(N)UM, TVNICA, SAGVM, P(A)ENVLA, PAL(L)A, PALLIOLUM, LODIX, BANATA}, and \emph{ABOLLA}, appear regularly enough without being abbreviated and
thus the interpretation of common abbreviations like \emph{l, la, lan, pan, t, sag, paenv, pal, lo, lod, lodi, bana, ab} is
not in doubt. The other abbreviations are mostly related to terms of colour. The prices on those tags are an important
information: they indicated the value of the goods or the cost of a given service (e.g. cleaning, fulling, dyeing). It
would also appear that they were used by fullers and dyers as ownership tags. By noting the name of the client as well
as the type of cloth or service and the price on the tag which was subsequently attached to the item to be processed
(i.e. cleaned or dyed), shop owners could easily return their property to the clients as well as charge them the
correct fee\footnote{\citet{radman_livaja_segestica_2008}, \citet{radman-livaja_roetiquettes_2010}, \citet{radman-livaja_5_2013}, \citet[165-172]{radman-livaja_two_2013}, \citet{radman-livaja_plombs_2014},  for analogous tags with \citet[97-104]{mocsy_olom_1956}; \citet[195-210]{egger_funf_1968}; \citet[127-138]{frei-stolba_les_2011}; \citet[121-137/93, 215-222]{schwinden_romerzeitliche_1985}; \citet{romer-martijnse_romerzeitliche_1990}; \citet[5-48]{romer-martijnse_fruhkaiserzeitliche_1997}; \citet[301-305]{feugere_etiquette_1993}; \citet[211-220]{weiss_bleietiketten_1991}; \citet[29-40]{paci_etichette_1995}; \citet[207-216]{bassi_tre_1996}; \citet[121-135]{bizzarini_quattro_2005}; \citet[43-51]{buchi_etichette_2005}; \citet[42-110]{cresci_supellex_2010}; \citet[295-317]{jacques_artisanat_2010}; \citet[237-246]{wedenig_bleietikett_2013}.}.

@@

\noindent In fact, the transcription and interpretation of Roman inscribed lead tags are often challenging, and so we ignored how
far we could go with these students if useable results would ensue. First we made a selection of fourty tags to be
subjected to the adequate digitization process we chose to work with: Reflectance Transforming Imaging (RTI). It is a
small but nevertheless representative sample of the whole corpus. The students had at their disposal different kinds of
inscriptions, including both easily readable specimens and more challenging graffiti (e.g., No. 12582). One may also
profit from the assistance of undergraduate and graduate students, which could not often be the case
before\footnote{\ The Laboratorio di Cultura Digitale, lead by Prof. Salvatori, at the University of Pisa applies the
didactic model DIGICRAFT. This means that undergraduate students might share their skills, especially the digital one,
in order to allow an entire team to reach its goal. For more information see: \url{http://www.labcd.unipi.it/laboratorio}.}.
This experiment showed that their contribution can be useful, because technology may compensate for their lack of
knowledge although the final conclusions have to remain in the domain of scholars familiar with the subject. In any
case, this experiment shows how the digital edition may allow both scholars and students to have an open access to what
is basically a primary source of information for humanities (presented as a digital facsimile)\footnote{\citet{gorman_opening_2015}
writes about some of the OA issues: ``While digitisation is not a pre-requisite to gaining access to material (...), and
while digital surrogates of cultural heritage objects do not have to be openly shared once created, just as the
sciences are calling for publication of source data as part of the Open Access movement, opening up access to primary
sources in the cultural heritage sector and encouraging them to be published in a way which is as accessible as
possible has the potential to change the nature of research outputs in the Humanities and Social Sciences, as well as
the nature of research itself in these areas.''}. Of course, learners could train themselves on the drawing of the tag,
but they would not confront the difficulty of interpreting the numerous ambiguities and the responsibility of deciding
what is on the tag.

@@

The drawing process consists in associating some heterogeneous elements of the wSystem and the tSystem. MarkOut is the
linking tool that allows to represent the graphico-textual (g/t) relationship between those two systems. MarkOut allows
the students to work on the wSystem of the inscriptions by drawing the signs over an image of the tag using the mouse.
The line is converted in a curve that can be later fixed using handles. Each mark, that represents a phenomenon of the
wSystem of the inscription, is assigned to some textual information such as a letter or a symbol (Unicode and/or XML
encoding), that represents a textual phenomenon of the tSystem of the inscription. This assignation creates inside the
MarkOut the g/t relationship used for query processing. The final result is an SVG file that can be easily parsed to
recover the linking information between wSystem and tSystem (e.g. browsing by glyph's shape and interpretation,
identifying glyph's position), and easily shared with anyone with a browser (e.g.: fig. \ref{fig:markout}). The linking tool MarkOut
is one of the editing components of the modular DAP. They fit into communication workflows\footnote{This raises, of
course, questions, most of them outside the purview of this article, about the relationship that evolves between
research and our modern societies. We invite the reader to peruse various initiatives such as \citet{bertocci_documenter_2006} on the «
valeur d'ancienneté » of cultural heritage, that participates to social well-being, the Civic Epistemologies
\url{http://www.civic-epistemologies.eu} and its Berlin charter, as well as Public History approaches
\url{http://public-history-weekly.oldenbourg-verlag.de}.} such as crowdsourcing, editing inscriptions, dialoguing with
experts, preserving and promoting cultural heritage and, as here, teaching some skills in epigraphy.

Fig. \ref{fig:markout}

\begin{figure}[!bp]
\centering
\includegraphics[scale=0.10]{EAGLE16lameetalteaching-img003.png}
\includegraphics[scale=0.10]{EAGLE16lameetalteaching-img004.png}
\caption{MarkOut screenshots (detailed part of tag no. 12563) showing the association, expressed in red, of the Unicode
Latin characters with the position and the shape of the writing: U+0045 with II (identified as E2 by students) and
U+004C with L (L1 or L5).}
\label{fig:markout}
\end{figure}

Students were involved for a very short period of time: on average two hours per day over a couple of days only (in
Spring 2014 and in Fall 2015). This training was delivered in three steps (``Readings'', ``Writings'', ``Test \& Taste a
DAP'') during which there were given some academic documentation about the lead tags, and basic knowledge about digital
epigraphy (epigraphic message in a computer, dispositive analysis, existing projects, digital tools, XML and Unicode
encoding, DAP, and issues regarding digital representation of artifacts).

Students encountered their first epigraphic challenges with tag no. 12582 \citep[cat 03.13, 347]{radman-livaja_plombs_2014}, found in
the Kupa river in Sisak and offered by Andrija Colussi to the Archaeological Museum in 1898. It is a rather typical
lead tag of an irregular rectangular form, pierced with a hole. Its size (22.7x22.8x2.4 mm) corresponds fairly well to
the average size of other specimens found in Siscia. Although its surface is rather damaged and shows clear traces of
erasure, the most recent inscription remains quite readable. However, many traces of an older record (or several older
records?), complicate significantly the transcription and the interpretation. The last inscription can be read,
nevertheless, but it demands a certain effort and quite a lot of experience in Latin palaeography:

\begin{minipage}[t]{0.25\textwidth}
\textbf{Obverse}\\
At\d{e}\d{r}\\
ivs
\end{minipage}
\begin{minipage}[t]{0.25\textwidth}
(palimpsest)\\
\d{S}\d{i}\d{i}\d{x}\d{t}\d{i}\\
i onis n\\
i . n i 
\end{minipage}
\begin{minipage}[c]{0.30\textwidth}
%\begin{figure}
%\centering
\includegraphics[width=2.965cm,height=1.79cm]{EAGLE16lameetalteaching-img005a.png}
%\end{figure}
\end{minipage}

\begin{minipage}[t]{0.25\textwidth}
\textbf{Reverse}\\
la(na) p(ondo) i\\
cor(ticea)\\
\sout{X} =-\pounds\\
\end{minipage}
\begin{minipage}[t]{0.25\textwidth}
(palimpsest)\\
. . . s\\
. . v\\
  iir 
\end{minipage}
\begin{minipage}[c]{0.30\textwidth}
%\begin{figure}
%\centering
\includegraphics[width=2.845cm,height=1.896cm]{EAGLE16lameetalteaching-img005b.png}
%\end{figure}
\end{minipage}

@@

@@

\begin{minipage}[t]{0.18\textwidth}
\textbf{Obverse}\\
Aeli(i)\\
Tasti
\end{minipage}
\begin{minipage}[c]{0.27\textwidth}
%\begin{figure}
%\centering
\includegraphics[width=2.965cm,height=1.79cm]{EAGLE16lameetalteaching-img006.png}
%\end{figure}
\end{minipage}
\begin{minipage}[t]{0.18\textwidth}
\textbf{Reverse}\\
\d{l}odicem\\
murtiola\d{m}\\
pondo vi s\\
\sout{X} vi s\\
\end{minipage}
\begin{minipage}[c]{0.25\textwidth}
%\begin{figure}
%\centering
\includegraphics[width=2.845cm,height=1.896cm]{EAGLE16lameetalteaching-img007.png}
%\end{figure}
\end{minipage}

No. 12563 is a rectangular lead tag of irregular form with a badly damaged and scratched surface. Nevertheless, the most
recent inscription is clearly readable while one may still discern traces of older inscriptions. An older inscription
might be visible on the obverse, perhaps series of numbers but its meaning remains unclear. Students have
transcriptions and drawings.

\begin{table}
{\small
\addtolength{\tabcolsep}{-0.7mm}
\begin{tabular*}{\textwidth}{llllllllll}
\toprule
\textbf{12563 (obv)} & \textbf{A} & \textbf{E} & \textbf{L} & \textbf{I} & \textbf{T} & \textbf{A} & \textbf{S} & \textbf{T} & \textbf{I}\\
\midrule
\textbf{Stud. 1} & \textit{A10} & \textit{E2} & \textit{L11} & \textit{I1} & \textit{T1} & \textit{A10} & \textit{S3} & \textit{T2} & \textit{I1} \\
\textbf{Stud. 2} & \textit{A10} & \textit{E2} & \textit{L5} & \textit{I1} & \textit{T2} & \textit{A10} & \textit{S4?} & \textit{T2} & \textit{I1}  \\
\bottomrule
\end{tabular*}

\begin{tabular*}{\textwidth}{llllllllllllll}
 & & & & & & & & & & & & & \\
\toprule
\textbf{12563 (rev)} & \textbf{L} & \textbf{O} & \textbf{D} & \textbf{I} & \textbf{C} & \textbf{E} & \textbf{M} & & & & & &\\
\midrule
\textbf{Stud. 1} & \textit{L2} & \textit{O3} & \textit{D3} & \textit{I1} & \textit{C2} & \textit{E2} & \textit{M5} &  &  & & & & \\
\textbf{Stud. 2} & \textit{?} & \textit{O3} & \textit{D5} & \textit{I2?} & \textit{C2} & \textit{E3} & \textit{M5} & & & & & &  \\
%\bottomrule
%\toprule
%& & & & & & & & & & & & & \\
\midrule
 & \textbf{M} & \textbf{U} & \textbf{R} & \textbf{T} & \textbf{I} & \textbf{O} & \textbf{L} & \textbf{A} & \textbf{M} & \textbf{P} & \textbf{V} & \textbf{I} & \textbf{S}\\
 \midrule
\textbf{Stud. 1}& \textit{M3} & \textit{V3} & \textit{R7} & \textit{T2} & \textit{I1} & \textit{O3} & \textit{L2} & \textit{A10} & \textit{M2} & \textit{P1} & \textit{V3} & \textit{I1} & \textit{S5} \\
\textbf{Stud. 2} & \textit{M3?} & \textit{V4} & \textit{R6} & \textit{T2} & \textit{I1} & \textit{O3} & \textit{L5} & \textit{A10} & \textit{M5} & \textit{P2} & \textit{V4} & \textit{I1} & \textit{S5} \\
\bottomrule
\end{tabular*}}

\caption{No. 12563.}
\label{tab:table1}
\end{table}

\begin{figure}[!bp]
\centering
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img009.png}
\caption{12563 Obverse. The two students' reconstructions}
\label{fig:1}
\end{figure}

\begin{figure}[!bp]
\centering
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img008.png}
\caption{12563 Reverse. The two students' reconstructions}
\label{fig:2}
\end{figure}

@@

%\begin{figure}
%\centering
%\includegraphics[width=2.009cm,height=1.96cm]{EAGLE16lameetalteaching-img010.png}
%\end{figure}
%\begin{figure}
%\centering
%\includegraphics[width=2.103cm,height=1.961cm]{EAGLE16lameetalteaching-img010.png}
%\end{figure}
%\begin{figure}
%\centering
%\includegraphics[width=2.044cm,height=1.794cm]{EAGLE16lameetalteaching-img011.png}
%\end{figure}

\begin{minipage}[t]{0.3\textwidth}
\textbf{Vale}\\
rius lis\\
Martia\\
\end{minipage}
\begin{minipage}[t]{0.3\textwidth}
(palimpsest)\\
\sout{X} \d{v}i\\
tiir \d{r} i\\
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img011a.png}
\end{minipage}

\begin{figure}[!bp]
\centering
\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img010.png}
\caption{13053 Obverse. Students' reconstructions}
\label{fig:3}
\end{figure}

@@

\begin{table}
{\small
\addtolength{\tabcolsep}{-0.8mm}
